### 2018-07-06 paper/5-implementations.tex
Give a bit more space to S5
Mostly whitespace tweaks

\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system; stores current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTimes) <- get
    let dirty = case Map.lookup key modTimes of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTimes)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Scheduler Applicative i i k v
topological rebuilder tasks target = execState $ mapM_ build order
  where
    build :: k -> State (Store i k v) ()
    build key = case tasks key of
        Nothing -> return ()
        Just task -> do
            store <- get
            let value = getValue key store
                newTask :: Task (MonadState i) k v
                newTask = rebuilder key value task
                fetch :: k -> State i v
                fetch k = return (getValue k store)
            newValue <- liftStore (run newTask fetch)
            modify $ putValue key newValue
    order = topSort (graph dep target)
    dep k = case tasks k of { Nothing -> []; Just task -> dependencies task }
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> [k] -- Throws error on a cyclic graph
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation
liftStore :: State i a -> State (Store i k v) a
liftStore x = do (a, newInfo) <- gets (runState x . getInfo)
                 modify (putInfo newInfo)
                 return a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-3mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty $ chain ++ [target | target `notElem` chain]
    modify $ mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:keys) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) keys
        Just task -> do
            store <- get
            let newTask :: Task (MonadState ir) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State ir (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            result <- liftStore (run newTask fetch) -- liftStore is defined in Fig. 7
            case result of
                Left dep -> go done $ dep : filter (/= dep) keys ++ [key]
                Right newValue -> do modify $ putValue key newValue
                                     (key :) <$> go (Set.insert key done) keys
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (implementation omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-5mm}
\end{figure}

@@

\vspace{-1mm}
\subsection{\Shake}\label{sec-implementation-shake}
\vspace{-1mm}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
        return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store = fst $ execState@\,@(fetch target)@\,@(store,@\,@Set.empty)
  where
    fetch :: k -> State (Store i k v, Set k) v
    fetch key = do
        done <- gets snd
        case tasks key of
            Just task | key `Set.notMember` done -> do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                newValue <- liftRun newTask fetch
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
                return newValue
            _ -> gets (getValue key . fst) -- fetch the existing value
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Run a task using a callback that operates on a larger state (implementation omitted)
liftRun :: Task (MonadState i) k v
        -> (k -> State (Store i k v, Set k) v) -> State (Store i k v, Set k) v
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-6mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    case cachedValues of
        _ | value `elem` cachedValues -> return value
        cachedValue:_ -> return cachedValue
        [] -> do (newValue, deps) <- track task fetch
                modify $ recordCT key newValue [ (k, hash v) | (k, v) <- deps ]
                return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to ctRebuilder
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
chain.}
\label{fig-cloud-implementations}
\vspace{-4mm}
\end{figure}

>>>>

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
\vspace{0.5mm}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system; stores current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTimes) <- get
    let dirty = case Map.lookup key modTimes of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTimes)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Scheduler Applicative i i k v
topological rebuilder tasks target = execState $ mapM_ build order
  where
    build :: k -> State (Store i k v) ()
    build key = case tasks key of
        Nothing -> return ()
        Just task -> do
            store <- get
            let value = getValue key store
                newTask :: Task (MonadState i) k v
                newTask = rebuilder key value task
                fetch :: k -> State i v
                fetch k = return (getValue k store)
            newValue <- liftStore (run newTask fetch)
            modify $ putValue key newValue
    order = topSort (graph dep target)
    dep k = case tasks k of { Nothing -> []; Just task -> dependencies task }
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> [k] -- Throws error on a cyclic graph
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation
liftStore :: State i a -> State (Store i k v) a
liftStore x = do
    (a, newInfo) <- gets (runState x . getInfo)
    modify (putInfo newInfo)
    return a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-5mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system; stores a dirty bit per key and calc chain
type Chain k = [k]
type ExcelInfo k = (k -> Bool, Chain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting dirtyBitRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on dirty bits
dirtyBitRebuilder :: Rebuilder Monad (k -> Bool) k v
dirtyBitRebuilder key value task = Task $ \fetch -> do
    isDirty <- get
    if isDirty key then run task fetch else return value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Scheduler Monad (ir, Chain k) ir k v
restarting rebuilder tasks target = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- liftChain $ go Set.empty $ chain ++ [target | target `notElem` chain]
    modify $ mapInfo $ \(ir, _) -> (ir, newChain)
  where
    go :: Set k -> Chain k -> State (Store ir k v) (Chain k)
    go _    []       = return []
    go done (key:keys) = case tasks key of
        Nothing -> (key :) <$> go (Set.insert key done) keys
        Just task -> do
            store <- get
            let newTask :: Task (MonadState ir) k (Either k v)
                newTask = try $ rebuilder key (getValue key store) task
                fetch :: k -> State ir (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            result <- liftStore (run newTask fetch) -- liftStore is defined in Fig. 7
            case result of
                Left dep -> go done $ dep : filter (/= dep) keys ++ [key]
                Right newValue -> do modify $ putValue key newValue
                                     (key :) <$> go (Set.insert key done) keys
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task that accepts a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (implementation omitted)
liftChain :: State (Store ir k v) a -> State (Store (ir, Chain [k]) k v) a
\end{minted}
\vspace{-2.5mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2.5mm}
\end{figure}

@@

\subsection{\Shake}\label{sec-implementation-shake}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system; stores verifying traces
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    upToDate <- verifyVT key (hash value) (fmap hash . fetch) =<< get
    if upToDate then return value else do
        (newValue, deps) <- track task fetch
        modify $ recordVT key (hash newValue) [ (k, hash v) | (k, v) <- deps ]
        return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Scheduler Monad i i k v
suspending rebuilder tasks target store = fst $ execState@\,@(fetch target)@\,@(store,@\,@Set.empty)
  where
    fetch :: k -> State (Store i k v, Set k) v
    fetch key = do
        done <- gets snd
        case tasks key of
            Just task | key `Set.notMember` done -> do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                newValue <- liftRun newTask fetch
                modify $ \(s, d) -> (putValue key newValue s, Set.insert key d)
                return newValue
            _ -> gets (getValue key . fst) -- fetch the existing value
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Run a task using a callback that operates on a larger state (implementation omitted)
liftRun :: Task (MonadState i) k v
        -> (k -> State (Store i k v, Set k) v) -> State (Store i k v, Set k) v
\end{minted}
\vspace{-1mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{-3mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    case cachedValues of
        _ | value `elem` cachedValues -> return value
        cachedValue:_ -> return cachedValue
        [] -> do (newValue, deps) <- track task fetch
                modify $ recordCT key newValue [ (k, hash v) | (k, v) <- deps ]
                return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to 'ctRebuilder'
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
chain.}
\label{fig-cloud-implementations}
\vspace{-2mm}
\end{figure}

### 2018-07-06 paper/6-engineering.tex
Mostly whitespace tweaks

One tempting way of modelling non-determinism is to enrich \hs{Tasks} from
\hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus},
respectively. More concretely, the following task description corresponds to
a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh3 :: Tasks MonadPlus String Integer
sprsh3 "B1"@\,@=@\,@Just $ Task $ \fetch -> (+) <$> fetch "A1" <*> (pure 1 <|> pure 2)
sprsh3 _   @\,@=@\,@Nothing
\end{minted}
\vspace{1mm}

@@

\begin{figure}
\vspace{-2mm}
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-build.pdf}}
\vspace{-1mm}
\caption{Initial build}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-clean.pdf}}
\vspace{-1mm}
\caption{Clean up and evict \cmd{main.prof}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-rebuild.pdf}}
\vspace{-1mm}
\caption{Build \cmd{main.prof} and \cmd{report.txt}}
\end{subfigure}
\vspace{-1mm}
\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
tool and record deep constructive traces in the cloud, (b)~remove built files
locally and evict \cmd{main.prof} from the cloud storage, (c)~build
\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
the new hash), then build \cmd{report.txt} by downloading it from the matching
deep constructive trace in the cloud, resulting in a Frankenbuild because
\cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted cloud
storage entries are highlighted; file hashes are shown in circles.
\label{fig-frankenbuild}}
\vspace{-4mm}
\end{figure}

Deep constructive traces (\S\ref{sec-deep-constructive-traces}) combined with
task non-determinism (\S\ref{sec-non-determinism}) can lead to very subtle bugs,
especially in the cloud setting. Fig.~\ref{fig-frankenbuild} shows a
\emph{Frankenbuild}~\cite{esfahani2016cloudbuild} example, where the target
\cmd{report.txt}, which is downloaded from the cloud, is inconsistent with its
immediate dependency \cmd{main.prof}. This inconsistency is caused by two factors:
(i)~inherent non-determinism of profiling: running a profiling tool on the very
same \cmd{main.exe} will produce different \cmd{main.prof} results every time;
and (ii)~relying on deep constructive traces, which cache build results based
only on the hashes of terminal task inputs (in this case \cmd{main.exe}). Note
that the resulting store is incorrect according to all three definitions of
correctness: the main definition (\S\ref{sec-build-correctness}), the variant
for non-deterministic tasks (\S\ref{sec-non-determinism}) and the variant for
shallow builds (this section).

>>>>

One tempting way of modelling non-determinism is to enrich \hs{Tasks} from
\hs{Applicative} or \hs{Monad} to \hs{Alternative} or \hs{MonadPlus},
respectively. More concretely, the following task description corresponds to
a spreadsheet with the formula \cmd{B1 = A1 + RANDBETWEEN(1,2)}.

\begin{minted}[xleftmargin=10pt]{haskell}
sprsh3 :: Tasks MonadPlus String Integer
sprsh3 "B1"@\,@=@\,@Just $ Task $ \fetch -> (+) <$> fetch "A1" <*> (pure@\,@1 <|> pure@\,@2)
sprsh3 _   @\,@=@\,@Nothing
\end{minted}
\vspace{1mm}

@@

\begin{figure}
\vspace{-2mm}
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-build.pdf}}
\caption{Initial build}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-clean.pdf}}
\caption{Clean up and evict \cmd{main.prof}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/frankenbuild-example-rebuild.pdf}}
\caption{Build \cmd{main.prof} and \cmd{report.txt}}
\end{subfigure}
\vspace{-1mm}
\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
tool and record deep constructive traces in the cloud, (b)~remove built files
locally and evict \cmd{main.prof} from the cloud storage, (c)~build
\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
the new hash), then build \cmd{report.txt} by downloading it from the matching
deep constructive trace in the cloud, resulting in a Frankenbuild because
\cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted cloud
storage entries are highlighted; file hashes are shown in circles.
\label{fig-frankenbuild}}
\vspace{-4mm}
\end{figure}

Deep constructive traces (\S\ref{sec-deep-constructive-traces}) combined with
task non-determinism (\S\ref{sec-non-determinism}) can lead to very subtle bugs,
especially in the cloud setting. Fig.~\ref{fig-frankenbuild} shows a
\emph{Frankenbuild}~\cite{esfahani2016cloudbuild} example, where the target
\cmd{report.txt}, which is downloaded from the cloud, is inconsistent with its
immediate dependency \cmd{main.prof}. This inconsistency is caused by two factors:
(i)~inherent non-determinism of profiling: running a profiling tool on the very
same \cmd{main.exe} will produce different \cmd{main.prof} results every time;
and (ii)~relying on deep constructive traces, which cache build results based
only on the hashes of terminal task inputs (in this case \cmd{main.exe}). Note
that the resulting store is incorrect according to all three definitions of
correctness: the main definition~(\S\ref{sec-build-correctness}), the variant
for non-deterministic tasks~(\S\ref{sec-non-determinism}) and the variant for
shallow builds~(this section).

### 2018-07-06 paper/7-related.tex
Mostly whitespace tweaks

As a simple example of using a build system for memoization, we solve a textbook
dynamic programming problem~--~Levenshtein's \emph{edit
distance}~\cite{levenshtein1966binary}: given two input strings $a$ and
$b$, find the shortest series of edit operations that transforms $a$
to $b$. The edit operations are typically \emph{inserting}, \emph{deleting} or
\emph{replacing} a symbol. The dynamic programming solution of this problem is
so widely known (e.g., see~\cite{cormen2001introduction}) that we provide its
encoding in our \hs{Task} abstraction without further explanation.
We address elements of strings $a_i$ and $b_i$ by keys \hs{A}~$i$ and \hs{B}~$i$,
respectively, while the cost of a subproblem $d_{ij}$ is identified by
\hs{D}~$i$~$j$.

\begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
data Key = A Integer | B Integer | D Integer Integer
\end{minted}
\vspace{0mm}
\begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
editDistance :: Tasks Monad Key Integer
editDistance (D i 0) = Just $ Task $ const $ pure i
editDistance (D 0 j) = Just $ Task $ const $ pure j
editDistance (D i j) = Just $ Task $ \fetch -> do
    ai <- fetch (A i)
    bj <- fetch (B j)
    if ai == bj then fetch (D (i - 1) (j - 1))
                else do insert  <- fetch (D  i      (j - 1))
                        delete  <- fetch (D (i - 1)  j     )
                        replace <- fetch (D (i - 1) (j - 1))
                        return (1 + minimum [insert, delete, replace])
editDistance _ = Nothing
\end{minted}

\noindent
When asked to build \hs{D}~$n$~$m$, a minimal build system will calculate the
result using memoization. Furthermore, when an input symbol $a_i$ is changed,
only necessary, incremental recomputation will be performed~--~an optimisation
that cannot be achieved just with memoization.

>>>>

As a simple example of using a build system for memoization, we solve a textbook
dynamic programming problem~--~Levenshtein's \emph{edit
distance}~\cite{levenshtein1966binary}: given two input strings $a$ and
$b$, find the shortest series of edit operations that transforms $a$
to $b$. The edit operations are typically \emph{inserting}, \emph{deleting} or
\emph{replacing} a symbol. The dynamic programming solution of this problem is
so widely known (e.g., see~\cite{cormen2001introduction}) that we provide its
encoding in our \hs{Task} abstraction without further explanation.
We address elements of strings $a_i$ and $b_i$ by keys \hs{A}~$i$ and \hs{B}~$i$,
respectively, while the cost of a subproblem $c_{ij}$ is identified by
\hs{C}~$i$~$j$.

\begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
data Key = A Int | B Int | C Int Int
\end{minted}
\vspace{0mm}
\begin{minted}[xleftmargin=10pt, fontsize=\small]{haskell}
editDistance :: Tasks Monad Key Int
editDistance (C i 0) = Just $ Task $ const $ pure i
editDistance (C 0 j) = Just $ Task $ const $ pure j
editDistance (C i j) = Just $ Task $ \fetch -> do
    ai <- fetch (A i)
    bj <- fetch (B j)
    if ai == bj
        then fetch (C (i - 1) (j - 1))
        else do insert  <- fetch (C  i      (j - 1))
                delete  <- fetch (C (i - 1)  j     )
                replace <- fetch (C (i - 1) (j - 1))
                return (1 + minimum [insert, delete, replace])
editDistance _ = Nothing
\end{minted}

\noindent
When asked to build \hs{C}~$n$~$m$, a minimal build system will calculate the
result using memoization. Furthermore, when an input symbol $a_i$ is changed,
only necessary, incremental recomputation will be performed~--~an optimisation
that cannot be achieved just with memoization.
